<style type="text/css">
.main-container {
max-width: 1800px;
margin-left: auto;
margin-right: auto;
}
</style>

---
title: "Paired data analysis adonis"
output: html_document
author: M.Davids
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

"Work in progress"

When dealing with longitudinal microbiome data, the subject confounding patters do not allow for regular adonis significance testing.
With a well designed permatution matrix it can still be used.

Data used is from RED study (https://www.sciencedirect.com/science/article/abs/pii/S1262363619301855).

# Install and load required packages

```{r environment, include=T, message=F, warning=F}
cran <- c("ggplot2",
          "ggpubr",
          "vegan",
          "permute",
          NULL)
install.packages(cran[!cran %in% installed.packages()])

bioc. <- c("phyloseq",
           "mixOmics",
           NULL)
BiocManager::install(bioc.[!bioc. %in% installed.packages()], update = F)

library(phyloseq)
library(mixOmics)
library(ggplot2)
library(ggpubr)
library(permute)
library(vegan)
```

# Read & wrangle data

```{r data, include=T, message=F, warning=F}
ps <- readRDS(url("https://github.com/AMCMC/MiCA-Tutorials/raw/master/Data_files/ps.2018_27_RED_study.2019-03-04.filtered.RDS"))
ps@sam_data$Treatment <- ps@sam_data$Randomization

ps <- phyloseq(otu_table(ps@otu_table@.Data[order(ps@sam_data$Subject_ID, ps@sam_data$Time_Point),], taxa_are_rows = F), tax_table(ps), sample_data(ps), ps@phy_tree)
```

Normalize the data by rarefaction without replacement.

```{r rarefy, include=T, warning=F, echo=T, message=F}
ps.rare<- rarefy_even_depth(ps, sample.size = 25000, rngseed = 21122, replace = F)
```

Only use subjects which have at least 2 time points. 

```{r subset, include=T}
ps.paired <- prune_samples(ps.rare@sam_data$Subject_ID %in%  names(which(table(ps.rare@sam_data$Subject_ID)!=1)), ps.rare)
table(ps.paired@sam_data$Treatment, ps.paired@sam_data$Time_Point)
```

The dataset contains 24 subjects that received DAPA, and 17 that recieved GLIC.

# Different adonis permutation models

Incorrect model with subject confounding

(Lets reorder the df such that all paired samples are together.)

```{r}
set.seed((1))
dist <- distance(ps.paired, "bray")
#dist <- distance(ps.paired, "wunifrac")
#dist <- distance(microbiome::transform(ps.paired, "clr"), method = "euclidean")

df <- data.frame(ps.paired@sam_data)
perm <- how(nperm = 999)
adonis2 (dist ~ Treatment, permutations = perm, data = df)
```

According to this test, treatment allocation is significantly associated with the microbiome.
This is misleading as subjects is confounded in the design.
A possible solution is to test the treatment time interaction.

```{r}
set.seed((1))
adonis2 (dist ~ Treatment*Time_Point, permutations = perm, data = df)
```

Here we see that overall the two treatment arms are different, independent of time, indicative of the (subject) bias. But are the two groups truly different?

With the following permutation design we can permute the time label within a subject and the treatment among the subjects (the correct design).

Lets have a look at the permutation matrix. 

```{r}
perm.strata <- how(plots = Plots(strata = df$Subject_ID, type = "free"))
set.seed(3)
perm <- shuffleSet(n=sample_names(ps.paired), control = perm.strata, nset = 3)
perm[,1:2]

df[1:2,c("Subject_ID","Treatment","Time_Point")]
droplevels(df$Subject_ID[t(perm[,1:2])])
droplevels(df$Treatment[t(perm[,1:2])])
droplevels(df$Time_Point[t(perm[,1:2])])
```

As you can see the samples are permuted in pairs, and the order of the pairs is shuffled.

```{r}
adonis2(dist ~ Treatment*Time_Point, permutations = perm.strata, data=df)
```

Since we now permuted time within a subject and treatment among subjects, we no longer have the subject confounding the design and we see that the two groups are actually not significantly different.

In conclusion we show there is no significant effect of the two diabetic drugs on the beta-diversity, in terms of Bray-Curtis dissimilarity, nor is there an allocation or time effect!

```{r}
sessionInfo()
````