<style type="text/css">
.main-container {
max-width: 1800px;
margin-left: auto;
margin-right: auto;
}
</style>

---
title: "Multi-level pca analysis"
output: html_document
author: M.Davids
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Multi-level PCA (mPCA) can be used to visualize longitudinal data were we want to remove the inter-individual variance, which can obscure possible treatment effects. The following code shows an example on a toy dataset. Once we understand how the data is tranformed, we can more easily understand the multi-level pca.

An alternative approach would be to use RDA. # add tutorial

# Install and load required packages

```{r environment, include=T, message=F, warning=F}
#renv::init()

cran <- c("ggplot2",
          "ggpubr",
          "ape",
          "ggfortify",
          NULL)

install.packages(cran[!cran %in% installed.packages()])

bioc. <- c("phyloseq",
           "mixOmics",
           NULL)
#BiocManager::install(bioc.[!bioc. %in% installed.packages()])

library(phyloseq)
library(mixOmics)
library(ggplot2)
library(ggpubr)
library(ggfortify)
```

# Toy data

Lets create a toy data set with 8 subjects and 16 features and two time points.

First we generate the baseline data. There is no confounding factor and all measurements are independent from the same distribution.

```{r}
nsample=8
nfeatures=16

set.seed(1);toy_data <- matrix(data = sample(x = 1:10, size = nsample*nfeatures, replace = T), nrow = nsample, ncol = nfeatures, dimnames = list(paste0("s",1:nsample,"_t1"),paste0("f",1:nfeatures)))

toy_data
```

Now lets generate the second time point data, which is completely independent of the baseline.

```{r}
set.seed(2);toy_data2 <- matrix(data = sample(x = 1:10, size = nsample*nfeatures, replace = T), nrow = nsample, ncol = nfeatures, dimnames = list(paste0("s",1:nsample,"_t2"),paste0("f",1:nfeatures)))

toy_data_merged <- rbind(toy_data, toy_data2)
toy_data_merged
```

Lets add a design

```{r}
toy_data_design <- data.frame(row.names = rownames(toy_data_merged),
                              treatment=factor(rep(c(rep(1,4),rep(2,4)),2)),
                              time_point=factor(c(rep(1,8),rep(2,8))),
                              subject=gsub("_.*","",rownames(toy_data_merged)))
toy_data_design
```

We can visualize using simple pca.

```{r}
pca_out <- prcomp(toy_data_merged)
autoplot(pca_out, data = toy_data_design, colour="treatment", shape="time_point") + 
  theme_bw() +
  geom_line(aes(group=subject)) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

As you can see all data is scattered and no obvious patterns, as it should
Now use mPCA to visualize the results.

```{r}
mpca <- pca(X = toy_data_merged, multilevel = toy_data_design$subject, ncomp = 10)

df <- cbind(mpca$variates, toy_data_design)

ggplot(df, aes(x=X.PC1, y=X.PC2)) +
  theme_bw() + 
  geom_point(aes(color=treatment, shape=time_point)) +
  geom_line(aes(group=subject), alpha=0.3) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

Also in this plot we dont see any patterns.

# Subject variance

Now lets make the second timepoint dependent on the first measurement and add some random noise by either substracting or adding one count, or not changing at all.

```{r data, include=T, message=T, warning=F}
set.seed(2);toy_data2 <- toy_data-sample(c(-1,0,1), size = nsample*nfeatures, replace = T)

toy_data_merged <- rbind(toy_data, toy_data2)
toy_data_merged
```

```{r}
pca_out <- prcomp(toy_data_merged)
autoplot(pca_out, data = toy_data_design, colour="treatment", shape="time_point") + 
  theme_bw() +
  geom_line(aes(group=subject)) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

Now we can see a clearly see all the inter-individual variance.
We can correct for all subject specific variance using mPCA.

```{r}
mpca <- pca(X = toy_data_merged, multilevel = toy_data_design$subject, ncomp = 10)

df <- cbind(mpca$variates, toy_data_design)

ggplot(df, aes(x=X.PC1, y=X.PC2)) +
  theme_bw() + 
  geom_point(aes(color=treatment, shape=time_point)) +
  geom_line(aes(group=subject), alpha=0.3) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

Now all interindivual subject variance has been removed from the ordination.
And no treatment or time effect can be observed.

Lets have a look at the transformed data that the mixOmics package generated.

```{r}
mpca$X
```

We either added or substracted 1 from the count table. As we can see this difference is averaged over the two time points. So effectively we are doing a pca on the deltas. This is the reason the plot is symmetrical. 

# Time effect

Now lets add some time effect by either adding or substracting.

```{r}
set.seed(3);toy_data3 <- toy_data2+matrix(rep(sample(c(-1,0,0,0,1), size = nfeatures, replace = T), nsample), ncol = nfeatures, byrow = T)

toy_data_merged <- rbind(toy_data, toy_data3)
toy_data_merged
```

```{r}
pca_out <- prcomp(toy_data_merged)
autoplot(pca_out, data = toy_data_design, colour="treatment", shape="time_point") + 
  theme_bw() +
  geom_line(aes(group=subject)) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

The interindividual differences are this the major source of variance and a small time effect can be observed (PC2).

```{r}
mpca <- pca(X = toy_data_merged, multilevel = toy_data_design$subject, ncomp = 10)

df <- cbind(mpca$variates, toy_data_design)

ggplot(df, aes(x=X.PC1, y=X.PC2)) +
  theme_bw() + 
  geom_point(aes(color=treatment, shape=time_point)) +
  geom_line(aes(group=subject), alpha=0.3) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

With the multilevel pca, time effect is very obvious.

# Treatment 

Now lets add treatment effect.

```{r}
set.seed(3);effect <- matrix(rep(sample(c(-1,0,1), size = nfeatures, replace = T), nsample), ncol = nfeatures, byrow = T)*2
effect[1:4,] <- 0

toy_data4 <- toy_data3+effect

toy_data_merged <- rbind(toy_data, toy_data4)
toy_data_merged
```


```{r}
pca_out <- prcomp(toy_data_merged)
autoplot(pca_out, data = toy_data_design, colour="treatment", shape="time_point") + 
  theme_bw() +
  geom_line(aes(group=subject)) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

There is no obvious treatment effect in the regular pca

```{r}
mpca <- pca(X = toy_data_merged, multilevel = toy_data_design$subject, ncomp = 10)

df <- cbind(mpca$variates, toy_data_design)

ggplot(df, aes(x=X.PC1, y=X.PC2)) +
  theme_bw() + 
  geom_point(aes(color=treatment, shape=time_point)) +
  geom_line(aes(group=subject), alpha=0.3) +
  stat_ellipse(aes(group=interaction(treatment, time_point)))
```

Now we can see an obvious effect of time and treatment.

# Summary

This is of course an exaggerated example, but it shows how inter-individual variance can obscure treatment effects. Especially in microbiome data, inter-individual variance is usally largerer than treatment effects.


```{r}
sessionInfo()
````